---
title: "Hwy 61 Overstory"
author: "Randy Swaty"
date: "2023-09-29"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## The basics--Hwy 61 Sites

Most data cleaning was in an Excel pivot table. Blanks were removed.
This should be considered draft subject to QA/QC. Raw data that includes
blanks is at bottom of page.

**As of September 28, 2023 species lists updated as recommended by Laura
Slavsky:**

Oaks - red oak, bur oak, swamp white oak

Non-planted - red maple, sugar maple, balsam fir, black ash, white ash,
paper birch

Aspen - quaking aspen, bigtooth aspen, balsam poplar

*All planted species represented individually*

**Filtered data to only contain HWY 61 Sites**

* Day Hill Fill
* Day Hill Gami
* Day Hill Gitchi
* Day Hill South
* Spilt Rock Rocky Top
* Split Rock 2021
* Split Rock 2021 NE Patch
* Split Rock Crazy Bay
* Split Rock Crazy East
* Split Rock Exclosure
* Split Rock Jon
* Split Rock Merrill 61
* Split Rock Pine
* Split Rock South



<br>

```{r set up, message=FALSE, warning=FALSE, include=FALSE}

# load packages

library(crosstalk)
library(ggsci)
library(kableExtra)
library(plotly)
library(scales)
library(tidyverse)


# read in data, round numbers to whole numbers
overstory_data_raw <- read_csv("data/overstory.csv") %>%
  mutate(ave_basal_area_whole = round(ave_basal_area, 0)) %>%
  group_by(site_name) %>%
  mutate(site_name = fct(rev(site_name)))
  
## remove blanks
overstory_data <- read_csv("data/overstory.csv") %>%
  mutate(ave_basal_area_whole = round(ave_basal_area, 0)) %>%
  group_by(site_name) %>%
  mutate(site_name = fct(rev(site_name))) %>%
  filter(!species_trees %in% "(blank)")

# update species list according to notes supplied by Laura Slavsky; removal of Howard Gnesen
overstory_data_updated<- overstory_data %>%
  mutate(updated_species = case_when(
    species_trees %in% c("BO", "RO", "SWO") ~ "Oaks",
    species_trees %in% c("RM", "SM", "BF", "BA", "WA", "PB") ~ "Non-planted",
    species_trees %in% c("QA", "BT", "BP") ~ "Aspen",
    TRUE ~ species_trees)) %>%
  filter(site_name !="Howard Gnesen")


hwy61_overstory <- overstory_data_updated %>%
  filter(site_name %in% c(
  "Day Hill Fill",
  "Day Hill Gami",
  "Day Hill Gitchi",
  "Day Hill South",
  "Spilt Rock Rocky Top",
  "Split Rock 2021",
  "Split Rock 2021 NE Patch",
  "Split Rock Crazy Bay",
  "Split Rock Crazy East",
  "Split Rock Exclosure",
  "Split Rock Jon",
  "Split Rock Merrill 61",
  "Split Rock Pine",
  "Split Rock South"
))

```

```{r grouped  by species, echo=FALSE, message=FALSE, warning=FALSE}

grouped_species <- hwy61_overstory %>%
  group_by(updated_species)%>%
  summarise(mean_basal_area = mean(ave_basal_area_whole))

grouped_species_chart  <- 
  ggplot(data = grouped_species, aes(x = updated_species, y = mean_basal_area)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(
    title = "Mean Basal Area by Species across all HWY 61 sites",
    x = "",
    y = "Basal Area") +
  scale_x_discrete(limits = rev(grouped_species$updated_species)) +
  coord_flip() +
  theme_bw()

grouped_species_chart
```

<br>

## Overstory basal area for HWY 61 Sites

```{r clickable species by site, echo=FALSE, message=FALSE, warning=FALSE}

# set up key for crosstalk

sd_overstory_data <- highlight_key(hwy61_overstory)



basal_area_chart <- 
  ggplot(sd_overstory_data, aes(x = updated_species,
                                 y = ave_basal_area_whole)) +
  geom_point(size = 3,
             colour = "#4c0185") +
  labs(
    title = "Average Basal Area per Species",
    x = "",
    y = "Basal Area") +
  scale_y_continuous(labels = comma) +
  theme_bw()

bscols(widths = c(3, 10),
       filter_select("BP", 
                     "Select site", 
                     sd_overstory_data, 
                     ~ site_name,
                     multiple = FALSE),
       ggplotly(basal_area_chart,
                tooltip = FALSE,
                width = 815)
)


```

<br>

## Looking across all of the HWY 61 sites

<br>

```{r all sites, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.width=9}
plot <-
  ggplot(hwy61_overstory, aes(fill = updated_species, y = ave_basal_area_whole, x = site_name)) +
  geom_bar(position = "stack", stat = "identity") +
  coord_flip() +
  labs(
    title = "Basal area per species by site",
    subtitle = "",
    caption = ".",
    x = "",
    y = "Basal Area, whatver that is :)",
    fill = "Status")+
  scale_x_discrete(limits=rev) +
  scale_y_continuous(label=comma) + 
  theme_bw() + 
  scale_fill_manual(values = c(
                              "#70504d", 
                              "#325ea8", 
                              "#0b4222",
                              "#6A3D9A", 
                              "#258a47", 
                              "#ab520e", 
                              "black",
                              "skyblue2", 
                              "#FB9A99", # lt pink
                              "red",
                              "#CAB2D6", # lt purple
                              "#FDBF6F", # lt orange
                              "gray70", 
                              "khaki2",
                              "maroon", 
                              "orchid1", 
                              "limegreen", 
                              "#ebf06e"))+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") 

#plot

ggplotly(plot)

```

<br>

## Pivoted and cleaned data for HWY 61 Sites

<br>

```{r data table, echo=FALSE, message=FALSE, warning=FALSE}

kable(hwy61_overstory,
      format.args = list(big.mark = ",")) %>%
  kable_styling(
    font_size = 12,
    bootstrap_options = c("striped", "hover", "condensed")
  ) 
```
