---
title: "Overstory Tree Data Exploration"
author: "Randy Swaty"
date: "2023-03-22"
output:
  html_document:
    toc: true
    toc_float: true
---


## The basics

Most data cleaning was in an Excel pivot table.  Blanks were removed.  This should be considered draft subject to QA/QC.  Raw data that includes blanks is at bottom of page. 

**As of September 28, 2023 species lists updated as recommended by Laura Slavsky:**

Oaks - red oak, bur oak, swamp white oak

Non-planted - red maple, sugar maple, balsam fir, black ash, white ash, paper birch

Aspen - quaking aspen, bigtooth aspen, balsam poplar

*All planted species represented individually*

<br>

```{r set up, message=FALSE, warning=FALSE, include=FALSE}

# load packages

library(crosstalk)
library(ggsci)
library(kableExtra)
library(plotly)
library(scales)
library(tidyverse)


# read in data, round numbers to whole numbers
overstory_data_raw <- read_csv("data/overstory.csv") %>%
  mutate(ave_basal_area_whole = round(ave_basal_area, 0)) %>%
  group_by(site_name) %>%
  mutate(site_name = fct(rev(site_name)))
  
## remove blanks
overstory_data <- read_csv("data/overstory.csv") %>%
  mutate(ave_basal_area_whole = round(ave_basal_area, 0)) %>%
  group_by(site_name) %>%
  mutate(site_name = fct(rev(site_name))) %>%
  filter(!species_trees %in% "(blank)")

# update species list according to notes supplied by Laura Slavsky
overstory_data_updated <- overstory_data %>%
  mutate(updated_species = case_when(
    species_trees %in% c("BO", "RO", "SWO") ~ "Oaks",
    species_trees %in% c("RM", "SM", "BF", "BA", "WA", "PB") ~ "Non-planted",
    species_trees %in% c("QA", "BT", "BP") ~ "Aspen",
    TRUE ~ species_trees
    
  ))
  

```




```{r grouped  by species, echo=FALSE, message=FALSE, warning=FALSE}

grouped_species <- overstory_data_updated %>%
  group_by(updated_species)%>%
  summarise(mean_basal_area = mean(ave_basal_area_whole))

grouped_species_chart  <- 
  ggplot(data = grouped_species, aes(x = updated_species, y = mean_basal_area)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(
    title = "Mean Basal Area by Species across all sites",
    x = "",
    y = "Basal Area") +
  scale_x_discrete(limits = rev(grouped_species$updated_species)) +
  coord_flip() +
  theme_bw()

grouped_species_chart
```

<br>

## Overstory basal area per site

```{r clickable species by site, echo=FALSE, message=FALSE, warning=FALSE}

# set up key for crosstalk

sd_overstory_data <- highlight_key(overstory_data_updated)



basal_area_chart <- 
  ggplot(sd_overstory_data, aes(x = updated_species,
                                 y = ave_basal_area_whole)) +
  geom_point(size = 3,
             colour = "#4c0185") +
  labs(
    title = "Average Basal Area per Species",
    x = "",
    y = "Basal Area") +
  scale_y_continuous(labels = comma) +
  theme_bw()

bscols(widths = c(3, 10),
       filter_select("BP", 
                     "Select site", 
                     sd_overstory_data, 
                     ~ site_name,
                     multiple = FALSE),
       ggplotly(basal_area_chart,
                tooltip = FALSE,
                width = 815)
)


```

<br>

## Looking across all of the sites

<br>


```{r all sites, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.width=9}
plot <-
  ggplot(overstory_data_updated, aes(fill = updated_species, y = ave_basal_area_whole, x = site_name)) +
  geom_bar(position = "stack", stat = "identity") +
  coord_flip() +
  labs(
    title = "Basal area per species by site",
    subtitle = "",
    caption = ".",
    x = "",
    y = "Basal Area, whatver that is :)",
    fill = "Status")+
  scale_x_discrete(limits=rev) +
  scale_y_continuous(label=comma) + 
  theme_bw() + 
  scale_fill_manual(values = c(
                              "#70504d", 
                              "#325ea8", 
                              "#0b4222",
                              "#6A3D9A", 
                              "#258a47", 
                              "#ab520e", 
                              "black",
                              "skyblue2", 
                              "#FB9A99", # lt pink
                              "red",
                              "#CAB2D6", # lt purple
                              "#FDBF6F", # lt orange
                              "gray70", 
                              "khaki2",
                              "maroon", 
                              "orchid1", 
                              "limegreen", 
                              "#ebf06e"))+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") 

#plot

ggplotly(plot)

```

<br>

## Pivoted and cleaned data (does have blanks; without updates)

<br>

```{r data table, echo=FALSE, message=FALSE, warning=FALSE}

kable(overstory_data_raw,
      format.args = list(big.mark = ",")) %>%
  kable_styling(
    font_size = 12,
    bootstrap_options = c("striped", "hover", "condensed")
  ) 
```





